The \emph{FrobeniusThresholds} package for \emph{Macaulay2}

### 2019-06-07 Fsing/FThresholdsPaper/FThresholds.tex
Edits to intro and nu section of paper
Work on Future Directions

This latter result is interesting from a computational perspective, in that it provides lower bounds for log canonical thresholds.  Furthermore, though there is a general purpose implementation of an algorithm for computing log canonical thresholds in the \emph{Dmodules} package by A.~Leykin and H.~Tsai \cite{DmodulesSource}, it happens that the function for computing $F$-pure thresholds contained in the \emph{FrobeniusThresholds} package is typically much faster, especially so in lower characteristic.  Note that the  \emph{MultiplierIdeals}  package  by Z.~Teitler, B.~Snapp, and C.~Raicu also computes log canonical thresholds in many special cases, including monomial ideals, hyperplane arrangements, generic determinantal ideals, and certain binomial ideals \cite{MultiplierIdealsPackage, MultiplierIdealsArticle}.

@@

\subsection{Some background and notation}
It is important to note that much of the functionality implemented in the \emph{FrobeniusThresholds} is not restricted to regular ambient rings; see Section \ref{sec.IsFPT} for a prominent example of this.

@@

Let us outline a common way in which natural numerical invariants in prime characteristic commutative algebra are constructed:  For every natural number $e$, associate to some fixed set of data|often, a collection of polynomials and ideals|an integer describing something of relevance that depends on $e$ (e.g., the dimension of some interesting vector space constructed in terms of your initial data).   Finally, normalize this integer by dividing by some power of $p^e$, and then take the limit as the integer $e$ tends to infinity.  The resulting limit, if it exists, should then encode some interesting information about your initial set of data.

Consider a nonzero polynomial $f$ and a natural number $e$.  If $f$ does not vanish at the origin, then set set $\nu_f^{\idealm}(p^e) := \infty$.  Otherwise, $f \in \idealm$, and we instead define
\[ \nu_f^{\idealm}(p^e) := \max \{ n>0 : f^n \notin \idealm^{[p^e]} \} \]
where $\idealm^{[p^e]}$ denotes the $p^e$-th Frobenius power of $\idealm$, that is, the ideal generated by the $p^e$-th powers of the elements of $\idealm$.    It is not very hard to see that this number is exactly the largest power of $f$ that is part of a free basis for the localization of the ambient polynomial ring at the origin, over its subring of $p^e$-th powers.  As noted earlier, this number also has a natural interpretation in terms of $F$-purity of pairs.

Following our outline, we define
\[ \ft^{\idealm}(f) := \lim_{e \to \infty} \frac{ \nu_f^{\idealm}(p^e)}{p^e}. \]

@@

In the literature, the $F$-pure threshold $\ft^{\idealm}(f)$ is often denoted $\fpt(f)$, for obvious reasons.  However, in this note, we adopt the former notation to avoid any possible confusion with the \emph{Macaulay2} function {\tt fpt} described in Section \ref{sec.FPT}.  Indeed, the command {\tt fpt(f)} need not always output the number $\ft^{\idealm}(f) = \fpt(f)$, though in the case that it cannot, it will provide a reasonably good estimate for this number.

@@

We conclude this subsection by briefly reviewing some natural generalizations.  Suppose that $I$ and $J$ are ideals.  If $I$ is contained in the radical of $J$, then we set
%
\[ \nu_I^J(p^e) = \max \{ n : I^n \notin J^{[p^e]} \}. \]
Otherwise, we set $\nu_I^J(p^e) = \infty$.
This clearly generalizes the quantity $\nu_f^{\idealm}(p^e)$ considered earlier, and we call
\[ \ft^J(I) := \lim_{e \to \infty} \frac{ \nu_I^J(p^e)}{p^e} \]
the \emph{$F$-threshold of $I$ with respect to $J$}.  These numbers are also rational, when they are finite.
The value $\ft^{\idealm}(I)$ is called the \emph{$F$-pure threshold of $I$}, and if $I = (f)$ is principal,  $\ft^{J}(f) := \ft^J(I)$ is called the \emph{$F$-threshold of $f$ with respect to $J$}.
Like $F$-pure thresholds, $F$-thresholds can also be defined in terms of its test ideals.

Finally, we point out that the  \emph{FrobeniusThresholds} package builds heavily upon the \emph{TestIdeals} package for \emph{Macaulay2} \cite{TestIdealsPackage, TestIdealsPaper,M2}, which provides a broad range of functionality for effective computation in prime characteristic commutative algebra.  In fact, both this package and \emph{TestIdeals} are derived (and have greatly grown) from an older package \emph{PosChar}.

@@

There are three main sections in this paper, which describe and illustrate the main functionality implemented in the package \emph{FrobeniusThresholds}.
\begin{itemize}
\item[$\circ$] \autoref{sec.Nu} describes the fundamental function {\tt nu}.  As its name suggests, this function is concerned with computing the integers $\nu_I^J(p^e)$ mentioned above.
\item[$\circ$] \autoref{sec.IsFPT} explains the functionality for determining whether a given number is the $F$-pure threshold, or more generally, an \emph{$F$-jumping exponent}.
\item[$\circ$] \autoref{sec.FPT} describes the features of the central function of the package, {\tt fpt}.
\item[$\circ$]  \autoref{sec.FutureDirections} outlines our plans for future improvements to this package.
\end{itemize}
%{\color{red} Move this later}

\subsection*{Acknowledgements.}  First and foremost, we enthusiastically thank everyone who helped complete the \emph{FrobeniusThresholds} package. 
Our package coauthors are Juliette Bruce and Daniel Smolkin.  Erin Bela, Zhibek Kadyrsizova, Moty Katzman, Sara Malec, and Marcus Robinson also contributed valuable code to the package.

@@

\subsection{The option {\tt UseSpecialAlgorithms} for the {\tt fpt} function}

@@

\newpage
\section{Future Directions}
\label{sec.FutureDirections}

We conclude with some comments on possible improvements to the package.

One natural candidate for improvement would be to extend functionality to better handle nonprincipal ideals.  In particular, the package could be improved if it could use the test ideals $\tau(I^t)$ when $I$ is not necessarily principal.  Fortunately, the theoretical foundation for computing such ideals has already largely been worked out in the paper \cite{SchwedeTuckerTestIdealsOfNonPrincipal}. However, such an improvement would necessarily require an update to the \emph{TestIdeals} package.

\begin{enumerate}
\item Add more cases to {\tt UseSpecialAlgorithms}.
%\item Add the capability to compute non-local {\tt fpt} and non-local {\tt nu} via an option.
\item Write a function that will extract from this sequences $\{ \nu_f^{\idealm}(p^e) \}_{e=1}^{\infty}$ roots of Bernstein-Sato polynomials.  Especially, strata.

\item Extend binary forms functions to handle more general $F$-thresholds (that is, with respect to diagonal ideals).  There is one tricky theoretical things, we don't know when this process would stop.  If we can do this, then we can compute $F$-pure thresholds of binary forms in non-standard gradings.  Cite the paper by Daniel and Pedro \cite{HernandezTeixeiraFThresholdFunctions}.
\end{enumerate}

\newpage
\bibliographystyle{skalpha}
\bibliography{MainBib}

>>>>

This latter result is interesting from a computational perspective, in that it provides lower bounds for log canonical thresholds.  Furthermore, though there is a general purpose implementation of an algorithm for computing log canonical thresholds in the \emph{Dmodules} package by A.~Leykin and H.~Tsai \cite{DmodulesSource}, it happens that the function for computing $F$-pure thresholds contained in the \emph{FrobeniusThresholds} package is typically much faster, especially so in lower characteristic.  Note that the  \emph{MultiplierIdeals}  package  by Z.~Teitler, B.~Snapp and C.~Raicu also computes log canonical thresholds in many special cases, including monomial ideals, hyperplane arrangements, generic determinantal ideals, and certain binomial ideals \cite{MultiplierIdealsPackage, MultiplierIdealsArticle}.

@@

\subsection*{Some background and notation}
It is important to note that much of the functionality implemented in the \emph{FrobeniusThresholds} is not restricted to regular ambient rings; see Section \ref{sec.IsFPT} for a prominent example of this.

@@

Let us outline a common way in which natural numerical invariants in prime characteristic commutative algebra are constructed:  For every natural number $e$, associate to some fixed set of data|often, a collection of polynomials and ideals|an integer describing something of relevance that depends on $e$ (e.g., the dimension of some interesting vector space constructed in terms of your initial data).   Finally, normalize this integer by dividing by some power of $p^e$, and then take the limit as the integer $e$ tends to infinity.  The resulting limit, should it exist, should then encode some interesting information about your initial set of data.

Consider a nonzero polynomial $f$ and a natural number $e$.  If $f$ does not vanish at the origin, then set set $\nu_f^{\idealm}(p^e) := \infty$.  Otherwise, $f \in \idealm$, and we instead define
\[ \nu_f^{\idealm}(p^e) := \max \{ n>0 : f^n \notin \idealm^{[p^e]} \} \]
where $\idealm^{[p^e]}$ denotes the $p^e$-th Frobenius power of $\idealm$, that is, the ideal generated by the $p^e$-th powers of the elements of $\idealm$.    It is not very hard to see that this number is exactly the largest power of $f$ that is part of a free basis for the localization of the ambient polynomial ring at the origin over its subring of $p^e$-th powers.  As noted earlier, this number also has a natural interpretation in terms of $F$-purity of pairs.

Following our outline, we define
\[ \ft^{\idealm}(f) := \lim_{e \to \infty} \frac{ \nu_f^{\idealm}(p^e)}{p^e} \]

@@

In the literature, the $F$-pure threshold $\ft^{\idealm}(f)$ is often denoted $\fpt(f)$, for obvious reasons.  However, in this note, we adopt the former notation to avoid any possible confusion with the \emph{Macaulay2} function {\tt fpt} described in Section \ref{sec.FPT}.  Indeed, the command {\tt fpt(f)} need not always output the number $\ft^{\idealm}(f) = \fpt(f)$, though in the case that it cannot, it will provide a reasonably good estimate of this number.

@@

We conclude this subsection by briefly reviewing some natural generalizations.  Suppose that $I$ and $J$ are ideals.  If $I$ is contained in the radical of $J$, then we set
%
\[ \nu_I^J(p^e) = \max \{ n : I^n \notin J^{[p^e]} \}. \]
Otherwise, we set $\nu_I^J(p^e) = \infty$.
This clearly generalizes the quantity $\nu_f^{\idealm}(p^e)$ considered earlier, and we call
\[ \ft^J(I) := \lim_{e \to \infty} \frac{ \nu_f^{\idealm}(p^e)}{p^e} \]
the \emph{$F$-threshold of $I$ with respect to $J$}.  These numbers are also rational, when they are finite.
The value $\ft^{\idealm}(I)$ is called the \emph{$F$-pure threshold of $I$}, and if $I = (f)$ is principal,  $\ft^{J}(f) := \ft^J(I)$ is called the \emph{$F$-threshold of $f$ with respect to $J$}.
Like $F$-pure thresholds, $F$-thresholds can also be defined in terms of its test ideals.

Finally, we point out that the  \emph{FrobeniusThresholds} package builds heavily upon the \emph{TestIdeals} package for \emph{Macaulay2} \cite{TestIdealsPackage, TestIdealsPaper,M2}, which provides a broad range of functionality for effective computation in prime characteristic commutative algebra.  In fact, both this package and \emph{TestIdeals} are derived from an older package \emph{PosChar}.

@@

There are three main sections in this paper, which describe and illustrate the main functionality implemented in the package \emph{FrobeniusThresholds}.
\begin{itemize}
\item[$\circ$] \autoref{sec.Nu} describes the fundamental function {\tt nu}.  As its name suggests, this function is concerned with computing the integers $\nu_I^J(p^e)$ described above.
\item[$\circ$] \autoref{sec.IsFPT} explains the functionality for determining whether a given number is the $F$-pure threshold, or more generally, an \emph{$F$-jumping exponent}.
\item[$\circ$] \autoref{sec.FPT} describes the features of the central function of the package, {\tt fpt}.
\item[$\circ$]  \autoref{sec.FutureDirections} describes our plans for future improvements to this package.
\end{itemize}
%{\color{red} Move this later}

\subsection*{Acknowledgements.}  First and foremost, the authors enthusiastically thank all authors and contributors to the
\emph{FrobeniusThresholds} package.  {\color{red} (CHECK THESE NAMES)}
Our package coauthors are Juliette Bruce and Daniel Smolkin.  Erin Bela, Zhibek Kadyrsizova, Moty Katzman, Sara Malec, and Marcus Robinson also contributed valuable code to the package.

@@

\subsection{The option {\tt UseSpecialAlgorithms}}

@@

\section{Possible Future Directions}
\label{sec.FutureDirections}

One natural future direction would be to better incorporate the test ideals $\tau(I^t)$ when computing $F$-thresholds in the case that the ideal $I$ is not necessarily principal.  The theoretical foundation for computing such ideals has already largely been worked out in the paper \cite{SchwedeTuckerTestIdealsOfNonPrincipal}, but such an update to the \emph{FrobeniusThresholds} package would likely the 
\emph{TestIdeals} package to be updated first.

At the moment, the specialized algorithm called by the {\tt fpt} function for computing the $F$-pure threshold of a homogeneous polynomial in two variables limits us to working with the standard grading on the ambient polynomial ring.  Clearly, it is desirable to be able to work with non-standard gradings as well, and results in the paper \cite{HernandezTeixeiraFThresholdFunctions} suggest an algorithm for doing so.  Thus, the package could be improved by finalizing, and implementing, such an algorithm.

Finally, it would be natural to to develop, and implement,  additional algorithms for computing $F$-pure threshold and $F$-jumping numbers in specialized situations apart from monomials, diagonals, binomials, and simple normal crossings.  The first, third, and fourth authors, along with Josep \'Alvarez Montaner and Jack Jeffries,  are currently working on developing such algorithms.  The theoretical foundation of these algorithms lies in polyhedral geometry and integer programming, and as such, should be able to be implemented in \emph{Macaualy2}.

\bibliographystyle{skalpha}
\bibliography{MainBib}
